---
title: "Geocoding: OpenStreetMap + R + Tableau"
subtitle: "AFFILIATION OR REMOVE" 
output:
  html_notebook: default
  html_document: default
---

-------------

## Notebook Instructions

-------------

* In many cases we have access to data with location information e.g. Address, City ,County ,Zip Code , Country.
To visualize this data with more detail we need to use geolocation.

* In order get the actual geolocation coordinates (Longitute,Latitude) for the location data, 
we need to calculate this coordinates.

* Using nominatim a free search engine for OpenStreetMap data to easily find out geolocation coordinates.

**OpenStreetMap:** Is open data built by a community of mappers that contribute and maintain 
data about roads, trails, cafés, railway stations, and much more, all over the world.

* https://www.openstreetmap.org/about

### Load Packages in R/RStudio 

We are going to use tidyverse a collection of R packages designed for data science. 
As well as other packages to help with data cleaning and processing. 

* tidiverse: https://www.tidyverse.org/
* rvest: https://github.com/tidyverse/rvest
* stringi: https://github.com/gagolews/stringi
* jsonlite: https://github.com/jeroen/jsonlite

-------------

## Geocoding: OpenStreetMap API Search Functions

-------------

The geocode function below allow us to make requests to Nominatim a search engine for OpenStreetMap data.

* Nominatim: https://nominatim.openstreetmap.org/
* Resources: https://nominatim.org/release-docs/develop/api/Search


*The chunk below checks that all the necessary packages are installed in the computer*

```{r, echo = FALSE, message=FALSE, warning=FALSE}

options(scipen = 9999)

# Here we are checking if the package is installed
if(!require("tidyverse")){
  # If the package is not in the system then it will be install
  install.packages("tidyverse", dependencies = TRUE)
  # Here we are loading the package
  library("tidyverse")
}

# Here we are checking if the package is installed
if(!require("rvest")){
  # If the package is not in the system then it will be install
  install.packages("rvest", dependencies = TRUE)
  # Here we are loading the package
  library("rvest")
}

# Here we are checking if the package is installed
if(!require("stringi")){
  # If the package is not in the system then it will be install
  install.packages("stringi", dependencies = TRUE)
  # Here we are loading the package
  library("stringi")
}

# Here we are checking if the package is installed
if(!require("lubridate")){
  # If the package is not in the system then it will be install
  install.packages("lubridate", dependencies = TRUE)
  # Here we are loading the package
  library("lubridate")
}


# Here we are checking if the package is installed
if(!require("jsonlite")){
  # If the package is not in the system then it will be install
  install.packages("jsonlite", dependencies = TRUE)
  # Here we are loading the package
  library("jsonlite")
}
```

```{r}
# INPUT LOCATION VARIABLES
# [Address], [City] , [State] , [Zipcode]

geocode <- function(record_id, address, city, state, zipcode){
  # NOMINATIM SEARCH API URL
  src_url <- "https://nominatim.openstreetmap.org/search?q="
  
  city <- str_replace_all(string = city, pattern = "\\s|,", replacement = "+")
  
  # CREATE A FULL ADDRESS
  addr <- paste(address, city, state, zipcode, sep = "%2C")
  # CREATE A SEARCH URL BASED ON NOMINATIM API TO RETURN GEOJSON
  requests <- paste0(src_url, addr, "&format=geojson")
  
  # ITERATE OVER THE URLS AND MAKE REQUEST TO THE SEARCH API
  for (i in 1:length(requests)) {
    
    # MAKE HTML REQUEST TO API AND TRANSFORME HTML RESPONSE TO JSON
    response <- read_html(requests[i]) %>%
      html_node("p") %>%
      html_text() %>%
      fromJSON()
    
    # FROM THE RESPONSE EXTRACT LATITUDE AND LONGITUDE COORDINATES
    lon <- response$features$geometry$coordinates[[1]][1]
    lat <- response$features$geometry$coordinates[[1]][2]
    
    # CREATE A COORDINATES DATAFRAME
    if(i == 1) {
      loc <- tibble(record_id = record_id[i], 
                    address = str_replace_all(addr[i], "%2C", ","),
                    latitude = lat, longitude = lon)
    }else{
      df <- tibble(record_id = record_id[i], 
                   address = str_replace_all(addr[i], "%2C", ","),
                   latitude = lat, longitude = lon)
      loc <- bind_rows(loc, df)
    }
  }
  return(loc)
}

```

-------------

## Data Source: San Francisco Buildings Permit Data

-------------

This data set pertains to all types of structural permits. Data includes details on application/permit numbers,
job addresses, supervisorial districts, and the current status of the applications.

* Source: https://data.sfgov.org/Housing-and-Buildings/Building-Permits/i98e-djp9

### Lets change the name of some columns and remote other that are irrelevant for this use case.

```{r}
data <- read_csv("https://data.sfgov.org/api/views/i98e-djp9/rows.csv?accessType=DOWNLOAD&bom=true&format=true")

View(data)

colremove = c("First Construction Document Date",
              "Structural Notification",
              "Number of Existing Stories",
              "Number of Proposed Stories",
              "Voluntary Soft-Story Retrofit",
              "Fire Only Permit","Existing Units",
              "Proposed Units","Plansets",
              "TIDF Compliance","Existing Construction Type",
              "Proposed Construction Type","Site Permit",
              "Supervisor District","Current Police Districts",
              "Current Supervisor Districts",
              "Current Status Date", "Permit Creation Date",
              "Analysis Neighborhoods","Lot","Location",
              "SF Find Neighborhoods","Unit","Block", "Permit Type",
              "Unit Suffix","Street Number Suffix",
              "Existing Construction Type Description")

data <- data[colnames(data)[1:47]] %>% select(-all_of(colremove))

data <- rename(data, "Neighborhoods Boundaries" = "Neighborhoods - Analysis Boundaries")

data <- rename(data, "Permit Type" = "Permit Type Definition")

```


### String Cleaning and Pattern Replacement

```{r}

bldg_df$city <- "San Francisco"
bldg_df$state <- "CA"
bldg_df$record_id <- as.character(bldg_df$record_id)

bldg_df$proposed_construction_type_description <-
  bldg_df$proposed_construction_type_description %>% 
  str_remove_all(pattern = "\\(|\\)") %>% 
  str_replace_all(" ", "_") %>% 
  as_factor()

bldg_df$street_name <- str_remove(string = bldg_df$street_name, pattern = "^0+")

bldg_df$street_name <- str_replace_all(string = bldg_df$street_name,
                                       pattern = "Bay Shore",
                                       replacement = "Bayshore")

bldg_df$street_suffix <- str_replace_all(string = bldg_df$street_suffix,
                                       pattern = "Bl",
                                       replacement = "Blvd")

bldg_df$street_suffix <- str_replace_all(string = bldg_df$street_suffix,
                                       pattern = "Tr",
                                       replacement = "Ter")

bldg_df$changed_use <- bldg_df$existing_use != bldg_df$proposed_use
bldg_df$estimated_revised_diff <- bldg_df$revised_cost - bldg_df$estimated_cost
```


### Date Types

```{r}
bldg_df$issued_date <- bldg_df$issued_date %>% mdy() %>% as_date()
bldg_df$issued_year <- bldg_df$issued_date %>% year()
bldg_df$issued_month <- bldg_df$issued_date %>% month(label = TRUE)
bldg_df$completed_date <- bldg_df$completed_date %>% mdy() %>% as_date()
bldg_df$completed_year <- bldg_df$completed_date %>% year()
bldg_df$completed_month <- bldg_df$completed_date %>% month(label = TRUE)

bldg_df$issued_completed_days <- bldg_df$completed_date - bldg_df$issued_date 
bldg_df$issued_completed_years <- bldg_df$completed_year - bldg_df$issued_year 

bldg_df$filed_date <- bldg_df$filed_date %>% mdy() %>% as_date()
bldg_df$filed_issued_days <- bldg_df$issued_date - bldg_df$filed_date

bldg_df$permit_expiration_date <- bldg_df$permit_expiration_date %>% mdy() %>% as_date()
bldg_df$issued_expiration_days <- bldg_df$permit_expiration_date - bldg_df$issued_date
```

### Project Cost and Zoning 

To look deeper into changes in zoning and type of constructions. We want to select only project that changed the building use. Also we will like to look at developments with significant capital investment in particular greater than half a million dollars 

```{r}

bldg_df <- bldg_df %>% arrange(desc(estimated_cost))
bldg_df <- bldg_df[bldg_df$changed_use == TRUE,]
bldg_df <- bldg_df %>% filter(revised_cost > 500000)

bldg_df$zipcode <- as_factor(bldg_df$zipcode)
bldg_df$permit_type <- as_factor(bldg_df$permit_type)
bldg_df$neighborhoods_boundaries <- as_factor(bldg_df$neighborhoods_boundaries)
bldg_df$proposed_use <- as_factor(bldg_df$proposed_use)

head(bldg_df)
```


### Preparing the address string variable to match the OpenStreeMaps API

```{r}
# ADDRESS VARIABLE MUST MATH NOMINATIM API
address <- paste0(bldg_df$street_number," ",
                  bldg_df$street_name," ",
                  bldg_df$street_suffix)

# DATA CLEANING SPECIAL CASES (e.g. 3333-3339 N CLARK)
query <- stri_replace(str = address, 
                      replacement = "", 
                      regex = "(-[0-9]+\\s)")

# REPLACE SPACES (\\s) OR COMMAS (,) WITH + SIGN
query <- str_replace_all(string = query, 
                         pattern = "\\s|,", 
                         replacement = "+")

head(query)
```


## Data Collection: OpenStreeMaps API Request 

```{r}
df <- geocode(record_id = bldg_df$record_id,
              address = query,
              city = bldg_df$city, 
              state = bldg_df$state,
              zipcode = bldg_df$zipcode)
```

Create a main dataset with all the relevant information

```{r echo = FALSE, message=FALSE, warning=FALSE}

bldg_df <- left_join(bldg_df, df, by="record_id")

bldg_df$address <- str_replace_all(string = bldg_df$address,
                                   pattern = "\\+",
                                   replacement = " ")

colremove <- c("street_number","street_name", 
                "street_suffix", "current_status")

bldg_df <- bldg_df %>% select(-one_of(colremove))

head(bldg_df)

```

Finally lets save the new dataset containing the restaurants Geolocation (latitude, longitude).

```{r}
write_csv(bldg_df, "data/building_geocoord.csv")
```
